---
title: "PCA"
output: html_document
date: "2024-07-17"
---

## IMPORT PACKAGES
```{r}
library(FactoMineR)
library(data.table)
library(ggplot2)
library(factoextra)
library(dendextend)
```

## IMPORT MATRIX
```{r}
matrix <- read.csv("/path/to/matrix.csv", header = TRUE, row.names = 1)
```

## IMPORT METADATA
```{r}
metadata <- read.csv("/path/to/metadata.csv", header = TRUE, row.names = 1, sep = ";")
```

## ADD INTEREST METADATA IN MATRIX
```{r}
matrix$col_name_metadata <- metadata$col_name_metadata[match(row.names(matrix), row.names(metadata))]
```

Strategy : 
1. CTRL + f (an element appears at the top of the code page)
2. Find = col_name_metadata and Replace = your metadata of interest (ex: group)
3. clic on All (at the end of the line)

## ACP
```{r}
matrix.acp <- PCA(matrix,graph=F, quali.sup = "col_name_metadata") 
```

## SCREEPLOT
```{r}
fviz_eig(matrix.acp, addlabels = TRUE, ylim = c(0, 60), barfill = "gray", barcolor = "gray") +
  ggtitle("Screeplot") +
  xlab("Main component") +
  ylab("Percentage of variance explained") +
  theme_classic() +  
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )
```

## GRAPH OF INDIVIDUALS
```{r}
plot(matrix.acp, habillage = "col_name_metadata", palette = c("human" = "blueviolet", "Cat" = "orange", "Monkey" = "yellowgreen", "Dog" = "blue"), cex = 0.5, title = "Individuals graph", label = "none")
```

## GRAPH OF VARIABLES
```{r}
plot(matrix.acp, choix = "var", cex = 0.5)
```

## CONTRIBUTION OF VARIABLES TO DIM 1
```{r}
nb_col <- ncol(matrix)
contrib <- matrix.acp$var$contrib * 100
sorted_contrib <- contrib[order(contrib[, "Dim.1"], decreasing = TRUE), ]
top_contrib <- sorted_contrib[1:nb_col-1, ]
top_contrib_df <- as.data.frame(top_contrib)
top_contrib_df$variable <- rownames(top_contrib_df)
ggplot(top_contrib_df, aes(x = reorder(variable, Dim.1), y = Dim.1)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top contributions of variables to dimension 1",
       x = "Variable",
       y = "Contribution to Dim.1 (%)") +
  theme_minimal() 
```

## CONTRIBUTION OF INDIVIDUALS TO DIM 1
```{r}
nb_row <- nrow(matrix)
contrib <- matrix.acp$ind$contrib 
sorted_contrib <- contrib[order(contrib[, "Dim.1"], decreasing = TRUE), ]
top_contrib <- sorted_contrib[1:nb_row-1, ]
top_contrib_df <- as.data.frame(top_contrib)
top_contrib_df$variable <- rownames(top_contrib_df)
top_contrib_df$col_name_metadata <- matrix$col_name_metadata[match(top_contrib_df$variable, rownames(matrix))]
ggplot(top_contrib_df, aes(x = reorder(variable, Dim.1), y = Dim.1, fill = col_name_metadata)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top contributions of individuals to dimension 1",
       x = "Individus",
       y = "Contribution to Dim.1 (%)") +
  theme_minimal() +
  scale_fill_manual(values = c("human" = "blueviolet", "Cat" = "orange", "Monkey" = "yellowgreen", "Dog" = "blue")) +
  theme(axis.text.y = element_text(size = 6))
```

## CLUSTERING 
```{r}
ind_coords <- matrix.acp$ind$coord
dist_matrix <- dist(ind_coords, method = "manhattan")
hc <- hclust(dist_matrix, method = "ward.D2")
dend <- as.dendrogram(hc)
group_colors <- as.character(matrix$col_name_metadata)
colors <- c("Human" = "blueviolet", "Cat" = "orange", "Monkey" = "yellowgreen", "Dog" = "blue")
group_colors <- colors[group_colors]
dend <- color_branches(dend, k = 4, groupLabels = FALSE)
labels_colors(dend) <- group_colors[order.dendrogram(dend)]
plot(dend, main = "Dendrogramme du clustering hiérarchique", cex = 0.5)
#rect.dendrogram(dend, k = 3, border = "red")
```

```{r}
inertie <- sort(hc$height, decreasing = TRUE)
plot(inertie[1:20], type = "s", xlab = "Number of classes", ylab = "Inertia", lwd = 2)
grid()
k <- 4 # to configure
abline(v = k, col = "red", lty = 3)
points(k, inertie[k], pch = 16, cex = 2, col = "red")
```





